Soundness Proofs for Iterative Deepening

### 2015-11-11 pldi-16/di.tex
Add narrative text about convergence; add proof sketch skeleton (lemmas unproved) for convergence

\subsection{Classifying data races as buggy or benign} % FIXME: dont say 'benign'; also, do you have enough left to say that this can even still be a full subsection?

@@

@@

Note especially that our reasoning does not require PPs on the internal locking of {\tt malloc} or {\tt free},
which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
Our full proof
{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
discusses our assumptions explicitly and provides the similar cases in more detail.
\renewcommand\proofname{Proof}

>>>>

@@

\subsection{Classifying data races as buggy or benign}

\subsection{Convergence to total verification}
\label{sec:totalverif}

Although Iterative Deepening's main purpose is to heuristically choose the most effective PP subsets to test
when the maximal state space is too large,
some tests may be small enough that their maximal state spaces can also be completed in time.
For such tests, a SSS-MC tool configured to preempt on every shared memory access \cite{spin} would provide a total verification of all possible thread schedules, if it could complete in time.
In this section, we show that Iterative Deepening provides a verification of the same strength if it completes the state spaces associated with every discovered data-race PP.
%In other words, contrapositively,
%if it is possible to find a bug with any sequence of preemptions on any instruction whatsoever,
%an equivalent thread interleaving will be reachable using only data-race PPs and synchronization API PPs.
A proof sketch of the contrapositive statement follows.

\begin{theorem}
If a bug can be exposed by any thread interleaving possible by preempting on any instruction,
Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
\end{theorem}
\begin{proof}
The proof has two parts:
first, we show that preempting on data-racing instructions and synchronization API boundaries suffices to test all possible program behavior;
second, we show that Iterative Deepening will eventually detect all such data races.

\begin{lemma}[Equivalence of irrelevant PPs]
For any thread interleaving possible by preempting on any instruction,
there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
	\label{lem:relevant}
\end{lemma}

asdfadsf % TODO

\begin{definition}[Reachable data race]
A data race candidate (or associated PP) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
\end{definition}
Initially, the statically-available sync API PPs are reachable. Reachability of data-race PPs is transitive.
\begin{lemma}[Saturation of data-race PPs]
	Given any interleaving comprising only data-race PPs and sync API PPs, all such PPs are reachable.
	\label{lem:saturation}
\end{lemma}

asdfasdfas % TODO

To conclude,
for any possible interleaving, Lemma \ref{lem:relevant} provides an equivalent one with only data-race and sync API PPs,
and Lemma \ref{lem:saturation} proves all involved PPs are reachable.
Hence, Iterative Deepening will eventually test a state space containing the equivalent buggy interleaving.
\end{proof}

Our full proof
{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
discusses our assumptions explicitly and includes more formal definitions and structure.

@@

Note especially that our reasoning does not require PPs on the internal locking of {\tt malloc} or {\tt free},
which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
Our full proof
{\em [see supplementary material]}
provides the similar cases in more detail.
\renewcommand\proofname{Proof}

### 2015-11-11 pldi-16/evaluation.tex
add structural text
Add narrative text about convergence; add proof sketch skeleton (lemmas unproved) for convergence

\section{Evaluation}

@@

\subsection{Finding nondeterministic data-race candidates}
\label{sec:eval-falseneg}
Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.
In such cases, a single-pass dynamic data-race detector
%could not achieve the coverage necessary
would fail
to identify a racing access pair as a candidate to begin with.
%
We counted how many such data-races, used as PPs, led to \quicksand~finding new bugs,
thereby making them {\em false negatives} of the single-pass approach.
% TODO: Put a figure here giving an example of where e.g. a data race only shows up during the contention path of a mutex.
We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).

@@

% TODO: Measure how much subset overlap there is.
{\bf Avoiding redundant work.}
When we extend a small state space with more PPs, the new state space is guaranteed to test a superset of thread interleavings compared to the old one.
Any interleaving which does not preempt threads on any of the new PPs will be repeated work.
Because we prioritize completing small state spaces before extending them with more PPs,
the superset state spaces we run later will always repeat each branch of their already-completed subsets.
%
We measured the proportion of repeated work among completed state spaces across our test suite;
on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
Predicting whether each still-untested interleaving will uncover new PPs is not straightforward,
but we believe future implementations of Iterative Deepening could incorporate cross-job memoization
to prune some or all of this repeated work.

>>>>

\section{Evaluation}
\label{sec:eval}

@@

% TODO: You should introduce this in the narrative as a key contribution way earlier.
\subsection{Finding nondeterministic data-race candidates}
\label{sec:eval-falseneg}
Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.
In such cases, a single-pass dynamic data-race detector
%could not achieve the coverage necessary
would fail
to identify a racing access pair as a candidate to begin with.
%
We counted how many such data-races, used as PPs, led to \quicksand~finding new bugs,
thereby making them {\em false negatives} of the single-pass approach.
% TODO: Put a figure here giving an example of where e.g. a data race only shows up during the contention path of a mutex.
We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).

@@

% TODO: Measure how much subset overlap there is.
{\bf Avoiding redundant work.}
When we extend a small state space with more PPs, the new state space is guaranteed to test a superset of thread interleavings compared to the old one.
Any interleaving which does not preempt threads on any of the new PPs will be repeated work.
Because we prioritize completing small state spaces before extending them with more PPs,
the superset state spaces we run later will always repeat each branch of their already-completed subsets.
%
We measured the proportion of repeated work among completed state spaces across our test suite;
on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.

In some cases, this repeated work can cause Iterative Deepening to be slower than SSS-MC to find certain bugs,
for example, if both {\tt mutex\_lock} and {\tt mutex\_lock} preemptions together would expose a bug, but not either alone.
Similarly, when pursuing total verification (\sect{\ref{sec:totalverif}}),
if the state space resulting from preempting on every instruction could be completed,
SSS-MC will achieve that verification faster, as Iterative Deepening will test many subsets of data-race PPs first.
Predicting whether each still-untested interleaving will uncover new PPs is not straightforward,
but we believe future implementations of Iterative Deepening could incorporate cross-job memoization
to prune some or all of this repeated work.

### 2015-11-11 pldi-16/intro.tex
Add narrative text about convergence; add proof sketch skeleton (lemmas unproved) for convergence

We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
Named after the analogous technique in chess AI \cite{iterative-deepening-chess-ai}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
Rather than attempting to search a single state space with every available preemption point enabled (e.g., preempting on every pthread API call),
\quicksand~tests many different state spaces corresponding to subsets of those points, managing a model checker instance to explore each one.
It estimates the size of each state space to decide when long-running instances should be suspended, and dynamically generates new state spaces based on data race analysis.

@@

This paper's contributions are as follows:
\begin{enumerate}
	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation of the technique;
	\item A new tactic for eliminating one class of false-positive data races,
		which cannot soundly be used in a single-pass analysis,
		but which we prove correct when used with Iterative Deepening;
		%, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
	%\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?
	\item A large evaluation in which \quicksand~compares favorably to stand-alone data-race detection and stateless model checking approaches.
\end{enumerate}

The remainder of the paper is organized as follows.
\sect{\ref{sec:design}} discusses the background and design of Iterative Deepening,
\sect{\ref{sec:implementation}} explains \quicksand's approach to implementing it,
\sect{\ref{sec:eval}} presents our evaluation,
\sect{\ref{sec:future}} discusses limitations and future work,
\sect{\ref{sec:related}} surveys the related work,
and \sect{\ref{sec:conclusion}} concludes.

>>>>

We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
Named after the analogous technique in chess AI \cite{iterative-deepening-chess-ai}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
Rather than attempting to search a single state space with every available preemption point enabled (e.g., preempting on every pthread API call),
\quicksand~tests many different state spaces corresponding to subsets of those points, managing a model checker instance to explore each one.
It estimates the size of each state space to decide when long-running instances should be suspended, and dynamically generates new state spaces based on data race analysis.
%In fact, if given enough time to fully test all discovered data-race preemption points,
%Iterative Deepening provides a full verification of all thread schedules that could arise from preempting anywhere.
In fact, Iterative Deepening is fully general:
we have proved that if it completes all state spaces resulting from data-race preemption points,
that serves as a total verification of all possible thread interleavings of the given test program.

@@

This paper's contributions are as follows:
\begin{enumerate}
	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation of the technique;
	\item A proof of convergence, which shows that fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules, should that be possible in the given CPU budget;
	\item A new tactic for eliminating one class of false-positive data races,
		which cannot soundly be used in a single-pass analysis,
		but which we prove correct when used with Iterative Deepening;
		%, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
	%\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?
	\item A large evaluation in which \quicksand~compares favorably to stand-alone data-race detection and stateless model checking approaches.
\end{enumerate}

The remainder of the paper is organized as follows.
\sect{\ref{sec:design}} discusses the background and design of Iterative Deepening, including our proof of convergence,
\sect{\ref{sec:implementation}} explains \quicksand's approach to implementing it, including our new false-positive data-race tactic,
\sect{\ref{sec:eval}} presents our evaluation,
\sect{\ref{sec:future}} discusses limitations and future work,
\sect{\ref{sec:related}} surveys the related work,
and \sect{\ref{sec:conclusion}} concludes.
